On Optimality of Finite Element Integration

### 2015-09-03 TOMS_Paper/assembly.tex
Incorporate Paul's advises and rework optimality

% Page heads
\markboth{F. Luporini et al.}{Optimal Finite Element Integration}

% Title portion
\title{Optimal Finite Element Integration}
\author{Fabio Luporini
\affil{Imperial College London}
David A. Ham
\affil{Imperial College London}
Paul H.J. Kelly
\affil{Imperial College London}}

\begin{abstract}
We tackle the problem of automatically generating optimal finite element integration routines for arbitrary multilinear forms. Optimality is defined in terms of floating point operations required to compute a result. We establish a model under which optimality can systematically be reached and comment on the general applicability of this model. The model is integrated in a compiler that, given a representation of an integration routine, derives a loop nest in optimal form. Extensive performance evaluation proves the robustness of the model, showing systematic performance improvements over a number of alternative code generation systems. The effect of low-level optimization is also discussed.
\end{abstract}

@@

\acmformat{Fabio Luporini, David A. Ham, and Paul   H. J. Kelly, 2015. Optimal Finite Element Integration.}

@@

The need for rapidly implementing high performance, robust, and portable finite element methods has led to approaches based on automated code generation. This has been proved successful in the context of the FEniCS (\cite{Fenics}) and Firedrake (\cite{firedrake-code}) projects, which have become increasingly popular over the last years. In these frameworks, the weak variational form of a problem is expressed at high-level by means of a domain-specific language. The mathematical specification is manipulated by a form compiler that generates a representation of assembly operators. By applying these operators to an element in the discretized domain, a local matrix and a local vector, which represent the contributions of that element to the equation solution, are computed. The code for assembly operators should be high performance: as the complexity of a variational form increases, in terms of number of derivatives, pre-multiplying functions, or polynomial order of the chosen function spaces, the operation count increases, thus making assembly covering a significant fraction of the overall runtime.

As demonstrated across a great deal of works, automating the generation of such high performance implementations poses several challenges. This is a result of the complexity inherent to the mathematical expressions involved in the numerical integration, which varies from problem to problem, and the particular structure of the loop nests enclosing the integrals. General-purpose compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent the expressions, thus producing sub-optimal code (i.e., code which performs more floating-point operations, or ``flops'', than necessary). Research compilers, for instance those based on polyhedral analysis of loop nests such as PLUTO (\cite{PLUTO}), focus on loop optimization for cache locality, so they are not particularly helpful in our context. The lack of suitable third-party tools has led to the development of a number of domain-specific code transformation (or synthesizer) systems. In~\citeN{quadrature1}, it is shown how automated code generation can be leveraged to introduce optimizations that a user should not be expected to write ``by hand''. In~\citeN{FFC-TC} and~\citeN{Francis}, mathematical reformulations of finite element integration are studied with the aim of minimizing the operation count. In~\citeN{Luporini}, the effects and the interplay of generalized code motion and a set of low-level optimizations are analysed. It is also worth mentioning an on-going effort to produce a novel form compiler, called UFLACS (\cite{Uflacs}), which adds to the already abundant set of code transformation systems for assembly operators.

However, in spite of such a considerable research effort, still there is no answer to one fundamental question: can we automatically generate an implementation of a form which is optimal in the number of flops executed? With this paper, we aim to make a first step towards solving this problem.

We summarize the contributions of this research as follows
\begin{itemize}
\item We characterize flop-optimality in a loop nest and we instantiate this concept to finite element integration. As part of this construction, we establish the notion of sharing and demonstrate that sharing can always be eradicated from the loop nests we are interested in.
\item We provide an incremental, constructive proof that optimal implementations can be obtain by analyzing the monomials appearing in a form
\item We automate the translation of forms into optimal loop nests by a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation. The compiler is open-source and available at \url{https://github.com/coneoproject/COFFEE}}, the default optimization system in the Firedrake framework.
\item We provide an unprecedented (for its extension) performance evaluation comparing several code transformation systems in a range of forms of increasing complexity. This is essential to demonstrate that our optimality model holds in practice.
\item We comment on the cases in which our characterization of optimality becomes weaker, and provide an intuition of how we handle such forms
\item We enrich the set of low-level transformations described in~\citeN{Luporini}. These make COFFEE a full-fledged system capable of decoupling form compilers from low level code generation.
\end{itemize}

@@

We consider the weak formulation of a linear variational problem
\begin{equation}
\begin{split}
Find\ u\ \in U\ such\ that \\
a(u, v) = L(v), \forall v \in V
\end{split}
\end{equation}
where $a$ and $L$ are, respectively, a bilinear and a linear form. The set of \textit{trial} functions $U$ and the set of \textit{test} functions $V$ are discrete function spaces. For simplicity, we assume $U = V$. Let $\lbrace \phi_i \rbrace$ be the set of basis functions spanning $U$. The unknown solution $u$ can be approximated as a linear combination of the basis functions $\lbrace \phi_i \rbrace$. From the solution of the following linear system it is possible to determine a set of coefficients to express $u$
\begin{equation}
A\textbf{u} = b
\end{equation}
in which $A$ and $b$ discretize $a$ and $L$ respectively:
\begin{equation}
\centering
\begin{split}
A_{ij} = a(\phi_i(x), \phi_j(x)) \\
b_i = L(\phi_i(x))
\end{split}
\end{equation}
The matrix $A$ and the vector $b$ are ``assembled'' and subsequently used to solve the linear system through (typically) an iterative method.

@@

\subsection{Quadrature Mode}
Quadrature schemes are typically used to numerically evaluate $A_{ij}^K$. For convenience, a reference element $K_0$ and an affine mapping $F_K : K_0 \rightarrow K$ to any element $K \in T$ are introduced. This implies a change of variables from reference coordinates $X_0$ to real coordinates $x = F_K (X_0)$ is necessary any time a new element is evaluated. The numerical integration routine based on quadrature over an element $K$ can be expressed as follows
\begin{equation}
\label{eq:quadrature}
A_{ij}^K = \sum_{q=1}^N \sum_{\alpha_3=1}^n \phi_{\alpha_3}(X^q)w_{\alpha_3} \sum_{\alpha_1=1}^d \sum_{\alpha_2=1}^d \sum_{\beta=1}^d \frac{\partial X_{\alpha_1}}{\partial x_{\beta}} \frac{\partial \phi_i^K(X^q)}{\partial X_{\alpha_1}} \frac{\partial X_{\alpha_2}}{\partial x_{\beta}} \frac{\partial \phi_j^K(X^q)}{\partial X_{\alpha_2}} det F_K' W^q
\end{equation}
where $N$ is the number of integration points, $W^q$ the quadrature weight at the integration point $X^q$, $d$ is the dimension of $\Omega$, $n$ the number of degrees of freedom associated to the local basis functions, and $det$ the determinant of the Jacobian matrix used for the aforementioned change of coordinates.

@@

\subsection{Qualitative Comparison}
\label{sec:qualitative}
Depending on form and discretization, the relative performance of the two modes, in terms of the operation count, can vary even quite dramatically. The presence of derivatives or coefficient functions in the input form tends to increase the size of the geometry tensor, making the traditional quadrature mode more indicate for ``complex'' forms. On the other hand, speed ups from adopting tensor mode can be significant in a wide class of forms in which the geometry tensor remains ``sufficiently small''. The discretization, particularly the relative polynomial order of trial, test, and coefficient functions, also plays a key role in the resulting operation count.

@@

\begin{Def}[Perfect and imperfect loop nests]
A loop nest is said to be \textit{perfect} when non-loop statements appear only in the body of the innermost loop. If this condition does not hold, a loop nest is said imperfect. 
\end{Def}

\begin{Def}[Independent basic block]
A sequence of statements that present no data dependencies compose an independent basic block
\end{Def}

@@

% In the following, we should represent loops by their iteration variable.
It is convenient to map mathematical properties to the loop nest level

@@

In particular, we are interested in the following class since it naturally arises from the math of finite element integration, reviewed in Section~\ref{sec:background}.

@@

\begin{Def}[Optimality of a multilinear loop nest]
The synthesis of a multilinear loop nest is \textit{optimal} if the amount of operations performed in the innermost loop is minimum.
\end{Def}

In other words, optimality implies there is no other synthesis able to further decrease the number of operations in the innermost loop. Note that this definition does not take into account memory requirements. If the loop nest were memory-bound -- the ratio of operations to bytes transferred from memory to the CPU being too low -- then speaking of ''optimality'' would clearly make no sense. In the following, we assume to operate in a CPU-bound regime, in which arithmetic-intensive expressions need be evaluated. This suits the context of finite element integration.

@@

\begin{Def}[Optimality of a loop nest]
\label{def:optimality}
The synthesis of a loop nest is \textit{optimal} if, under a set of memory constraints \texttt{C}, the total amount of operations performed in all innermost loops is minimum.
\end{Def}

Note how the definition contemplates the possibility for a nest to have multiple innermost loops. In fact, multiple sub-nests could be rooted in the outermost loop, either because part of the input or as a result of some transformations (e.g., see Figure~\ref{code:loopnest_nored}).

@@

We make the following observations. 1) $L >>> M, N, O$; that is, the number of elements $L$ is typically order of magnitude larger than both quadrature points ($M$) and degrees of freedom ($N$ and $O$ for test and trial functions); 2) $[L_j, L_k]$ (or simply $L_j$ with a linear form) is perfect and multilinear; this naturally descends from the translation of Equation~\ref{eq:quadrature} into a loop nest.

@@

\subsection{Memory constraints}
\label{sec:mem-const}
The fact that $L >>> M, N, O$ suggests we should be cautious about hoisting out of $LN$. Imagine $LN$ is enclosed in a time stepping loop. One could think of exposing and then hoisting time-invariant sub-expressions. There are several kinds of time-invariant sub-expressions: for example, those depending on the geometry (i.e., depending on $L_e$) and those which do not (e.g., when only involving reference element terms). In the former case, code motion increases the working set by a factor $L$ (unless adopting complex engineering solutions, like blocking non-affine loop nests, which are practically difficult to devise and maintain). The drop in number of operations could then be overwhelmed, from the runtime viewpoint, by a much larger memory pressure.

@@

\section{Code Generation}
\label{sec:codegen}
The analysis described in Section~\ref{sec:optimal-synthesis} has been fully automated in COFFEE, the optimization system for local assembly used in Firedrake. In this section, we describe the structure of the code generation system and we also comment on a set of low-level optimizations.

@@

We evaluate the speed ups achieved by a wide variety of transformation systems over the original code (i.e., no optimizations applied) as returned by the FEniCS Form Compiler. We analyze the impact of
\begin{itemize}
\item FEniCS Form Compiler: optimized quadrature mode (work presented in~\citeN{quadrature1}). Referred to as \texttt{quad} 
\item FEniCS Form Compiler: tensor mode (work presented in~\citeN{FFC-TC}). Referred to as \texttt{tens} 
\item FEniCS Form Compiler: automatic mode (choice between \texttt{tens} and \texttt{quad} driven by heuristic, detailed in~\citeN{Fenics} and summarized in Section~\ref{sec:qualitative}). Referred to as \texttt{auto} 
\item UFLACS: a novel back-end for the FEniCS Form Compiler (whose primary goals are improved code generation time and runtime). Referred to as \texttt{ufls} 
\item COFFEE: generalized loop-invariant code motion and padding (work presented in~\citeN{Luporini}). Referred to as \texttt{cfO1} 
\item COFFEE: optimal multilinear loop nest synthesis, padding and symbolic execution (captures the contributions of this paper). Referred to as \texttt{cfO2} 
\end{itemize}

@@

The upcoming Figures (\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity}) can be interpreted as ``nested'' plots. We refer to the outer plot as the ``grid'', while the inner are the actual ``plots''. In a grid, $p$ varies along the horizontal axis and $q$ varies along the vertical axis. The top-left plot in a grid shows the speed up over original code for $[q=1, p=1]$; the plot on its right for $[q=1, p=2]$, and so on. The diagonal of the grid provides the behaviour when test, trial and coefficient functions have same polynomial order, that is $q=p$. A grid can therefore be looked at from different perspectives, which allows us to make structured considerations on the performance achieved. In each plot there are six groups of bars, each group referring to a particular code variant (\texttt{quad, tens, ...}). There are four bars per group: the leftmost bar corresponds to the case $nf = 0$, the one on its right to the case $nf = 1$, and so on.

@@

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/mass}}
\caption{Performance evaluation for the mass matrix. The bars represent speed up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:mass}
\end{figure}

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/helmholtz}}
\caption{Performance evaluation for the bilinear form of a Helmholtz equation. The bars represent speed up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:helmholtz}
\end{figure}

\begin{figure}
\includegraphics[scale=0.7]{perf-results/elasticity}
\caption{Performance evaluation for the bilinear form arising in a elastic model. The bars represent speed up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:elasticity}
\end{figure}

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/hyperelasticity}}
\caption{Performance evaluation for the bilinear form arising in a hyperelastic model. The bars represent speed up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:hyperelasticity}
\end{figure}

@@

The first observation is that our optimality model does not imply minimum execution time. In particular, in the 0.06$\%$ of the test cases (we are not counting marginal differences), \texttt{cfO2} does not result in the best runtime. There are two reasons for this. The former is that low level optimization can have a significant impact. Compare, as an example, \texttt{tens} and \texttt{cfO2} in \texttt{Mass} and \texttt{Helmholtz} when $q=1$. With $p \in [2,3]$, \texttt{tens} slightly outperforms \texttt{cfO2}. With $p=4$, \texttt{tens} is quite faster than \texttt{cfO2}. The motivation is the aggressive unrolling performed in \texttt{tens}, which 1) allows to eliminate all operations involving zero-valued entries and 2) can drastically decrease both working set and register pressure by avoiding storing entire temporaries (there are often many repeated values that, with unrolling, need be allocated only once). In COFFEE we never unroll loops to maximize the chances of autovectorization, which usually results in considerable speed ups throughout the majority of test cases. The latter reason is simply inherent to our optimality model. As explained in Section~\ref{sec:heuristic-int-dep}, we use heuristics to optimize the integration loop. The reason we are studying an hyperelastic model is indeed to assess their effectiveness; the systematic performance improvements observed over all other variants are, however, extremely encouraging.

@@

Also note some other minor things.
\begin{itemize}
\item The lack of (or marginal) improvements in all code variants when $[q=1, p=1]$ is due to the cost of matrix insertion, which outweighs that of local assembly. 
\item The slow downs experienced with \texttt{quad} are imputable to the presence of indirection arrays in the generated code; especially with dense tables, trading vectorization for avoiding iteration over a few zero-valued columns may be counter-productive (see also Section~\ref{sec:zeros}).
\item A missing bar in \texttt{tens} or \texttt{auto} means that the code generation system failed because of either exceeding the memory limit or being unable to manipulate the math characterizing the form (this happens systematically in the hyperelastic model, and for some discretizations in the elastic model).
\item The bars for $nf=0$ do not change as $p$ increases: this is expected since varying $p$ when there are no coefficients should not make any difference.
 \end{itemize}

@@

\paragraph{Helmholtz}
Figure~\ref{fig:helmholtz} shows that the performance improvements achieved by various code variants over a non-optimized implementation can be significant. We appreciate the fact that the relative order of the employed function spaces plays a key role in the code synthesis choice. Take, for example, the case $[q=3, p=1, nf=2]$. The adoption of pre-evaluation makes a tremendous speed up be achievable by \texttt{cfO2} over the competitors. On the other hand, \texttt{auto} triggers \texttt{quad} (since one monomial in the form has two derivatives and two coefficients, which exceeds the threshold within which \texttt{tens} is selected), which, just like the other quadrature-based variants \texttt{ufls} and \texttt{cfO1}, is sub-optimal. The effects of the working set threshold are in general positive. Moving to $T_H = L3_{size}$ makes \texttt{cfO2} slower than \texttt{cfO1} in many cases with $nf=2$, the most significant one being a slow down of 2.16$\times$ at $[q=2, p=2]$.

\paragraph{Elasticity}
The performance results for the elastic model are displayed in Figure~\ref{fig:elasticity}. The fact that \texttt{auto} opts for \texttt{tens} when $nf=1$ leads, generally, to sub-optimal execution times. This is different than in the case of the Helmholtz equation, in which the choice of \texttt{tens} was generally correct when $nf=1$. Pre-evaluation is never triggered in \texttt{cfO2}, because of either being estimated sub-optimal or exceeding $T_H$. In the former case, the performance improvements exhibited by \texttt{cfO2} over \text{cfO1} (and, in general, over all code variants) are due to completely removing sharing and avoiding iteration over zero-valued blocks. The latter case occurs with $q=4$ and $nf \in [0, 1]$. Running the same experiments with $T_H = L3_{size}$ resulted in a considerable improvement when $nf = 0$, with \texttt{cfO2} becoming even faster than \texttt{tens}, while slow downs characterized the cases of $nf=1$ for all values of $p$. Our explanation for this is that when $[q=4, nf=0]$, $T_H$ is exceeded by only roughly 40$\%$, while the cost function predicts a save in operation count of 5$\times$. This suggests that our model could be refined to handle the cases in which the gain in operation count is so large to make non-extreme variations in working set size negligible.

>>>>

% Page heads
\markboth{F. Luporini et al.}{On Optimality of Finite Element Integration}

% Title portion
\title{On Optimality of Finite Element Integration}
\author{Fabio Luporini
\affil{Imperial College London}
David A. Ham
\affil{Imperial College London}
Paul H.J. Kelly
\affil{Imperial College London}}

\begin{abstract}
We tackle the problem of automatically generating optimal finite element integration routines given a high level specification of arbitrary multilinear forms. Optimality is defined in terms of floating point operations required to execute a loop nest. The generation of optimal loop nests is driven by a model that exploits mathematical properties of the domain of interest. A theoretical analysis and extensive experimentation prove the effectiveness of our approach, showing systematic performance improvements over a number of alternative code generation systems. The effect of low-level optimization is also discussed.
\end{abstract}

@@

\acmformat{Fabio Luporini, David A. Ham, and Paul   H. J. Kelly, 2015. On Optimality of Finite Element Integration.}

@@

The need for rapidly implementing high performance, robust, and portable finite element methods has led to approaches based on automated code generation. This has been proved successful in the context of the FEniCS (\cite{Fenics}) and Firedrake (\cite{firedrake-code}) projects, which have become increasingly popular over the last years. In these frameworks, the weak variational form of a problem is expressed at high-level by means of a domain-specific language. The mathematical specification is manipulated by a form compiler that generates a representation of assembly operators. By applying these operators to an element in the discretized domain, a local matrix and a local vector, which represent the contributions of that element to the equation solution, are computed. The code for assembly operators should be high performance: as the complexity of a variational form increases, in terms of number of derivatives, pre-multiplying functions, or polynomial order of the chosen function spaces, the operation count increases, with the result that assembly often accounts for a significant fraction of the overall runtime.

As demonstrated by the considerable body of research on the topic, automating the generation of such high performance implementations poses several challenges. This is a result of the complexity inherent to the mathematical expressions involved in the numerical integration, which varies from problem to problem, and the particular structure of the loop nests enclosing the integrals. General-purpose compilers, such as \emph{GNU's} and \emph{Intel's}, fail at exploiting the structure inherent in the expressions, thus producing sub-optimal code (i.e., code which performs more floating-point operations, or ``flops'', than necessary). Research compilers, for instance those based on polyhedral analysis of loop nests such as PLUTO (\cite{PLUTO}), focus on parallelization and loop optimization for cache locality, so they are not particularly helpful in our context. The lack of suitable third-party tools has led to the development of a number of domain-specific code transformation (or synthesizer) systems. In~\citeN{quadrature1}, it is shown how automated code generation can be leveraged to introduce optimizations that a user should not be expected to write ``by hand''. In~\citeN{FFC-TC} and~\citeN{Francis}, mathematical reformulations of finite element integration are studied with the aim of minimizing the operation count. In~\citeN{Luporini}, the effects and the interplay of generalized code motion and a set of low-level optimizations are analysed. It is also worth mentioning an on-going effort to produce a novel form compiler, called UFLACS (\cite{Uflacs}), which adds to the already abundant set of code transformation systems for assembly operators.

However, in spite of such a considerable research effort, still there is no answer to one fundamental question: can we automatically generate an implementation of a form which is optimal in the number of flops executed? In this paper, we formulate an approach to solve this problem. Summarizing, our contributions are as follows

\begin{itemize}
\item We characterize flop-optimality in a loop nest and we instantiate this concept to finite element integration. As part of this construction, we establish the notion of sharing and demonstrate that sharing can always be eradicated from the loop nests we are interested in.
\item We provide a model centred on sharing and other mathematical properties of our domain of interest; the model drives the translation of a monomial appearing in a form into flop-optimal loop nests.
\item We comment on the cases in which such model does not lead to optimal loop nests. We will see that this might occur with particular forms that make extensive use of tensor algebra. We show, however, that our model still is capable of producing quasi-optimal loop nests.
\item We integrate the model with a compiler, COFFEE\footnote{COFFEE stands for COmpiler For Fast Expression Evaluation. The compiler is open-source and available at \url{https://github.com/coneoproject/COFFEE}}, which is in use in the Firedrake framework.
\item We experimentally evaluate using a broader suite of forms, discretizations, and code generation systems than has been used in prior research. This is essential to demonstrate that our optimality model holds in practice.
\end{itemize}

In addition, in order to place COFFEE on the same level of other code generation systems from the viewpoint of low-level optimization (which is essential for a fair performance comparison)

\begin{itemize}
\item We introduce an engine based on symbolic execution that allows skipping irrelevant floating point operations (e.g., those involving zero-valued quantities). We elaborate on the performance impact of this optimization, making a clear distinction between flop optimality and efficient code.
\end{itemize}

@@

We consider the weak formulation of a linear variational problem
\begin{equation}
\begin{split}
Find\ u\ \in U\ such\ that \\
a(u, v) = L(v), \forall v \in V
\end{split}
\end{equation}
where $a$ and $L$ are, respectively, a bilinear and a linear form. The set of \textit{trial} functions $U$ and the set of \textit{test} functions $V$ are discrete function spaces. For simplicity, we assume $U = V$. Let $\lbrace \phi_i \rbrace$ be the set of basis functions spanning $U$. The unknown solution $u$ can be approximated as a linear combination of the basis functions $\lbrace \phi_i \rbrace$. From the solution of the following linear system it is possible to determine a set of coefficients to express $u$:
\begin{equation}
A\textbf{u} = b
\end{equation}
in which $A$ and $b$ discretize $a$ and $L$ respectively:
\begin{equation}
\centering
\begin{split}
A_{ij} = a(\phi_i(x), \phi_j(x)) \\
b_i = L(\phi_i(x))
\end{split}
\end{equation}
The matrix $A$ and the vector $b$ are ``assembled'' and subsequently used to solve the linear system through (typically) an iterative method.

@@

\subsection{Quadrature Mode}
Quadrature schemes are typically used to numerically evaluate $A_{ij}^K$. For convenience, a reference element $K_0$ and an affine mapping $F_K : K_0 \rightarrow K$ to any element $K \in T$ are introduced. This implies that a change of variables from reference coordinates $X_0$ to real coordinates $x = F_K (X_0)$ is necessary any time a new element is evaluated. The numerical integration routine based on quadrature over an element $K$ can be expressed as follows
\begin{equation}
\label{eq:quadrature}
A_{ij}^K = \sum_{q=1}^N \sum_{\alpha_3=1}^n \phi_{\alpha_3}(X^q)w_{\alpha_3} \sum_{\alpha_1=1}^d \sum_{\alpha_2=1}^d \sum_{\beta=1}^d \frac{\partial X_{\alpha_1}}{\partial x_{\beta}} \frac{\partial \phi_i^K(X^q)}{\partial X_{\alpha_1}} \frac{\partial X_{\alpha_2}}{\partial x_{\beta}} \frac{\partial \phi_j^K(X^q)}{\partial X_{\alpha_2}} det F_K' W^q
\end{equation}
where $N$ is the number of integration points, $W^q$ the quadrature weight at the integration point $X^q$, $d$ is the dimension of $\Omega$, $n$ the number of degrees of freedom associated to the local basis functions, and $det$ the determinant of the Jacobian matrix used for the aforementioned change of coordinates.

@@

\subsection{Qualitative Comparison}
\label{sec:qualitative}
Depending on form and discretization, the relative performance of the two modes, in terms of the operation count, can vary quite dramatically. The presence of derivatives or coefficient functions in the input form tends to increase the size of the geometry tensor, making the traditional quadrature mode preferable for ``complex'' forms. On the other hand, speed-ups from adopting tensor mode can be significant in a wide class of forms in which the geometry tensor remains ``sufficiently small''. The discretization, particularly the relative polynomial order of trial, test, and coefficient functions, also plays a key role in the resulting operation count.

@@

\begin{Def}[Perfect and imperfect loop nests]
A perfect loop nest is a loop whose body either 1) comprises only a sequence of non-loop statements or 2) is itself a perfect loop nest. If this condition does not hold, a loop nest is said imperfect. 
\end{Def}

\begin{Def}[Independent basic block]
An independent basic block is a sequence of statements that present no data dependencies
\end{Def}

@@

% In the following, we should represent loops by their iteration variable.
It is convenient to map mathematical properties to the loop nest level.

@@

In particular, we are interested in the following class since it naturally arises in finite element integration (as reviewed in Section~\ref{sec:background}).

@@

\begin{Def}[Relaxed optimality of a multilinear loop nest]
The synthesis of a multilinear loop nest is optimal if the number of operations performed in the innermost loop is minimum.
\end{Def}

In other words, optimality implies there is no other synthesis able to further decrease the number of operations in the innermost loop. We use the attribute ``relaxed'' to denote that the definition does not take into account outer loops. In the following, we assume ``relaxed'' is implicit whenever we discuss optimality or mention ``optimal form''. Also note that this definition does not take into account memory requirements. If the loop nest were memory-bound -- the ratio of operations to bytes transferred from memory to the CPU being too low -- then speaking of optimality would clearly make no sense. In the following, we assume to operate in a CPU-bound regime, in which arithmetic-intensive expressions need be evaluated. This suits the context of finite element integration.

@@

\begin{Def}[Relaxed optimality of a loop nest]
\label{def:optimality}
The synthesis of a loop nest is optimal if, under a set of memory constraints \texttt{C}, the total amount of operations performed in all innermost loops is minimum.
\end{Def}

Note how the definition contemplates the possibility for a nest to have multiple innermost loops. In fact, multiple sub-nests could be rooted in the outermost loop, either because part of the input or as a result of some transformations (e.g., see Figure~\ref{code:loopnest_nored}). Again, we defined``relaxed optimality'' to highlight the fact that we are focussing on inner loops only.

@@

We make the following observations. 1) $L \gg M, N, O$; that is, the number of elements $L$ is typically order of magnitude larger than both quadrature points ($M$) and degrees of freedom ($N$ and $O$ for test and trial functions); 2) $[L_j, L_k]$ (or simply $L_j$ with a linear form) is perfect and multilinear; this naturally descends from the translation of Equation~\ref{eq:quadrature} into a loop nest.

@@

\subsection{Memory constraints}
\label{sec:mem-const}
The fact that $L \gg M, N, O$ suggests we should be cautious about hoisting out of $LN$. Imagine $LN$ is enclosed in a time stepping loop. One could think of exposing and then hoisting time-invariant sub-expressions. There are several kinds of time-invariant sub-expressions: for example, those depending on the geometry (i.e., depending on $L_e$) and those which do not (e.g., when only involving reference element terms). In the former case, code motion increases the working set by a factor $L$ (unless adopting complex engineering solutions, like blocking non-affine loop nests, which are practically difficult to devise and maintain). The drop in number of operations could then be overwhelmed, from the runtime viewpoint, by a much larger memory pressure.

@@

\section{Code Generation}
\label{sec:codegen}
The analysis described in Section~\ref{sec:optimal-synthesis} has been fully automated in COFFEE, the optimization system for local assembly used in Firedrake. In this section, we describe the structure of the code generation system.

@@

We evaluate the speed-ups achieved by a wide variety of transformation systems over the original code (i.e., no optimizations applied) as returned by the FEniCS Form Compiler. We analyze the impact of
\begin{itemize}
\item FEniCS Form Compiler: optimized quadrature mode (work presented in~\citeN{quadrature1}). Referred to as \texttt{quad} 
\item FEniCS Form Compiler: tensor mode (work presented in~\citeN{FFC-TC}). Referred to as \texttt{tens} 
\item FEniCS Form Compiler: automatic mode (choice between \texttt{tens} and \texttt{quad} driven by heuristic, detailed in~\citeN{Fenics} and summarized in Section~\ref{sec:qualitative}). Referred to as \texttt{auto} 
\item UFLACS: a novel back-end for the FEniCS Form Compiler (whose primary goals are improved code generation time and runtime). Referred to as \texttt{ufls} 
\item COFFEE: generalized loop-invariant code motion and padding (work presented in~\citeN{Luporini}). Referred to as \texttt{cfO1} 
\item COFFEE: optimal multilinear loop nest synthesis, padding and symbolic execution (captures the contributions of this paper). Referred to as \texttt{cfO2} 
\end{itemize}

@@

The upcoming Figures (\ref{fig:mass},~\ref{fig:helmholtz},~\ref{fig:elasticity}, and~\ref{fig:hyperelasticity}) can be interpreted as ``nested'' plots. We refer to the outer plot as the ``grid'', while the inner are the actual ``plots''. In a grid, $p$ varies along the horizontal axis and $q$ varies along the vertical axis. The top-left plot in a grid shows the speed-up over original code for $[q=1, p=1]$; the plot on its right for $[q=1, p=2]$, and so on. The diagonal of the grid provides the behaviour when test, trial and coefficient functions have same polynomial order, that is $q=p$. A grid can therefore be looked at from different perspectives, which allows us to make structured considerations on the performance achieved. In each plot there are six groups of bars, each group referring to a particular code variant (\texttt{quad, tens, ...}). There are four bars per group: the leftmost bar corresponds to the case $nf = 0$, the one on its right to the case $nf = 1$, and so on.

@@

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/mass}}
\caption{Performance evaluation for the mass matrix. The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:mass}
\end{figure}

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/helmholtz}}
\caption{Performance evaluation for the bilinear form of a Helmholtz equation. The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:helmholtz}
\end{figure}

\begin{figure}
\includegraphics[scale=0.7]{perf-results/elasticity}
\caption{Performance evaluation for the bilinear form arising in a elastic model. The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:elasticity}
\end{figure}

\begin{figure}
 \makebox[\textwidth][c]{\includegraphics[scale=0.77]{perf-results/hyperelasticity}}
\caption{Performance evaluation for the bilinear form arising in a hyperelastic model. The bars represent speed-up over the original (unoptimized) code produced by the FEniCS Form Compiler.}\label{fig:hyperelasticity}
\end{figure}

@@

The first observation is that our optimality model does not imply minimum execution time. In particular, in the 0.06$\%$ of the test cases (we are not counting marginal differences), \texttt{cfO2} does not result in the best runtime. There are two reasons for this. The former is that low level optimization can have a significant impact. Compare, as an example, \texttt{tens} and \texttt{cfO2} in \texttt{Mass} and \texttt{Helmholtz} when $q=1$. With $p \in [2,3]$, \texttt{tens} slightly outperforms \texttt{cfO2}. With $p=4$, \texttt{tens} is quite faster than \texttt{cfO2}. The motivation is the aggressive unrolling performed in \texttt{tens}, which 1) allows to eliminate all operations involving zero-valued entries and 2) can drastically decrease both working set and register pressure by avoiding storing entire temporaries (there are often many repeated values that, with unrolling, need be allocated only once). In COFFEE we never unroll loops to maximize the chances of autovectorization, which usually results in considerable speed-ups throughout the majority of test cases. The latter reason is simply inherent to our optimality model. As explained in Section~\ref{sec:heuristic-int-dep}, we use heuristics to optimize the integration loop. The reason we are studying an hyperelastic model is indeed to assess their effectiveness; the systematic performance improvements observed over all other variants are, however, extremely encouraging.

@@

Also note some other minor things.
\begin{itemize}
\item The lack of (or marginal) improvements in all code variants when $[q=1, p=1]$ is due to the cost of matrix insertion, which outweighs that of local assembly. 
\item The slow-downs experienced with \texttt{quad} are imputable to the presence of indirection arrays in the generated code; especially with dense tables, trading vectorization for avoiding iteration over a few zero-valued columns may be counter-productive (see also Section~\ref{sec:zeros}).
\item A missing bar in \texttt{tens} or \texttt{auto} means that the code generation system failed because of either exceeding the memory limit or being unable to manipulate the math characterizing the form (this happens systematically in the hyperelastic model, and for some discretizations in the elastic model).
\item The bars for $nf=0$ do not change as $p$ increases: this is expected since varying $p$ when there are no coefficients should not make any difference.
 \end{itemize}

@@

\paragraph{Helmholtz}
Figure~\ref{fig:helmholtz} shows that the performance improvements achieved by various code variants over a non-optimized implementation can be significant. We appreciate the fact that the relative order of the employed function spaces plays a key role in the code synthesis choice. Take, for example, the case $[q=3, p=1, nf=2]$. The adoption of pre-evaluation makes a tremendous speed-up be achievable by \texttt{cfO2} over the competitors. On the other hand, \texttt{auto} triggers \texttt{quad} (since one monomial in the form has two derivatives and two coefficients, which exceeds the threshold within which \texttt{tens} is selected), which, just like the other quadrature-based variants \texttt{ufls} and \texttt{cfO1}, is sub-optimal. The effects of the working set threshold are in general positive. Moving to $T_H = L3_{size}$ makes \texttt{cfO2} slower than \texttt{cfO1} in many cases with $nf=2$, the most significant one being a slow-down of 2.16$\times$ at $[q=2, p=2]$.

\paragraph{Elasticity}
The performance results for the elastic model are displayed in Figure~\ref{fig:elasticity}. The fact that \texttt{auto} opts for \texttt{tens} when $nf=1$ leads, generally, to sub-optimal execution times. This is different than in the case of the Helmholtz equation, in which the choice of \texttt{tens} was generally correct when $nf=1$. Pre-evaluation is never triggered in \texttt{cfO2}, because of either being estimated sub-optimal or exceeding $T_H$. In the former case, the performance improvements exhibited by \texttt{cfO2} over \text{cfO1} (and, in general, over all code variants) are due to the elimination of sharing and the avoidance of iteration over zero-valued blocks. The latter case occurs with $q=4$ and $nf \in [0, 1]$. Running the same experiments with $T_H = L3_{size}$ resulted in a considerable improvement when $nf = 0$, with \texttt{cfO2} becoming even faster than \texttt{tens}, while slow-downs characterized the cases of $nf=1$ for all values of $p$. Our explanation for this is that when $[q=4, nf=0]$, $T_H$ is exceeded by only 40$\%$, while the cost function predicts a save in operation count of 5$\times$. This suggests that our model could be refined to handle the cases in which the gain in operation count is so large that non-extreme variations in working set size should be considered negligible.
